---
title: "3_Calculating_Watershed_Characteristics"
author: "Katie Willi"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(caret)
library(terra)
library(elevatr)
library(dataRetrieval)
library(nhdplusTools)
library(StreamCatTools)
library(tmap)
library(climateR)
library(data.table)
library(mapview)
list.files("src/", full.names = TRUE) %>% walk(~source(.))
```

Pull in our WSR gage watersheds:
```{r}
watersheds_inside <- list.files("data/nwis_ws_backup/inside", full.names = TRUE) %>%
  map(~readRDS(.)) %>%
  bind_rows() %>%
  rowid_to_column("index") %>%
  st_transform(4326)
```

Pull in our outside WSR gage watersheds:
```{r}
outside_list <- str_remove(list.files("data/nwis_data/outside"), "\\.csv$")

watersheds_outside <- st_read("data/nwis_ws_backup/outside/") %>% # also found at: "data/archive/shapefiles/watersheds.shp"
  filter(!site_no %in% watersheds_inside$site_no &
           site_no %in% outside_list) %>%
  left_join(., read_csv("data/nwis_data/all_outside_gages.csv"), by = "site_no") %>%
  select(site_no, comid) %>%
  st_transform(4326)
```

Bind together:
```{r}
watersheds <- bind_rows(watersheds_inside, watersheds_outside) %>%
  select(site_no, comid) %>%
  rowid_to_column(var = "index") %>%  
  .[1:2,]
```


##### STATSGO data from StreamCate

```{r}
selected_vars <- c(
  # Soil, STATSGO variables:
  "Clay", "Sand", "Rckdep", "WtDep", "Perm") %>% 
  tolower()

streamcat_sim <- watersheds %>%
  simple_streamcat_data(.) %>%
  mutate(across(where(is.numeric), ~ifelse(is.nan(.), 0, .))) %>%
  mutate(across(where(is.numeric), ~ifelse(. < 0, 0, .))) %>%
  st_drop_geometry() %>%
  rename_with(~ paste0(., "_streamcat"), -index) 
```

##### Aspect features

Dominant aspect requires a bit of a wonky workflow - I'm not quite sure if there's an easier approach than what I'm presenting here. I grab raw elevation DEMs, get the aspect (in degrees) for each grid cell, then convert those raw aspects (in degrees) into categorical N, E, S, W cardinal directions.

```{r}
watersheds_aspects <- watersheds %>%
  aspect_grabber(.) %>%
  st_drop_geometry() 
write_csv(watersheds_aspects, "data/watershed_aspects.csv")
```

##### Snow persistence

Here I am loading in all the snow persistence data I downloaded from: https://www.sciencebase.gov/catalog/item/5f63790982ce38aaa23a3930. There is annual snow persistence data from 2001-2020. Using the {terra} package, I get the area-weighted annual average snow persistence value for each watershed, then average each year's data together to get a single time- and area-weighted average for each watershed.

```{r}
watersheds_sp <- watersheds %>%
  snow_persistence_grabber(.) %>%
  select(-c(comid)) %>% 
  st_drop_geometry()
```

##### Daymet variables

```{r}
watersheds_daymet <- watersheds %>%
  daymet_grabber(.) %>%
  select(-c(comid)) %>% 
  st_drop_geometry() %>%
  select(-c(swe_avg_sum, swe_jul, swe_aug, swe_sep, swe_oct, swe_nov, swe_dec))
```


##### gridMET variables
```{r}
watersheds_pet <- watersheds %>%
  select(-site_no) %>%
  get_climate_historic_frac_overlap(., col_name = "index")

pet_avg <- watersheds_pet %>%
  as.data.table() %>%
  mutate(wyear = ifelse(month(date) >= 10, year(date) + 1, year(date))) %>%
  group_by(site_no, x, y, area_weight, wyear) %>%
  summarize(total_pet = sum(pet, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(site_no, x, y, area_weight) %>%
  summarize(mean_tot_pet = mean(total_pet)) %>%
  ungroup() %>%
  group_by(site_no) %>%
  summarize(
    avg_tot_pet = sum(mean_tot_pet * area_weight) / sum(area_weight),
    .groups = "drop"
  ) %>%
  rename(index = site_no)
```

##### Geology variables

```{r}
watersheds_geology <- watersheds %>%
  geology_grabber(.) %>%
  select(-c(comid)) %>% 
  st_drop_geometry()

# Vector of geology column names
geology_cols <- c(
  "perc_intrusive", "perc_metamorphic", "perc_volcanic",
  "perc_sedimentary_clastic", "perc_sedimentary_chemical",
  "perc_sedimentary_carbonate", "perc_unconsolidated", "perc_mix"
)

# create dominant_geology column
watersheds_geology <- watersheds_geology %>%
  rowwise() %>%
  mutate(dominant_geology = sub("^perc_", "", geology_cols[which.max(c_across(all_of(geology_cols)))])) %>%
  ungroup() %>%
  select(index, dominant_geology)
```

##### NLCD variables

```{r}
watersheds_nlcd <- watersheds %>%
  nlcd_grabber(.) %>%
  st_drop_geometry() %>%
  select(index, Forest_Percent, Grassland_Percent,
         Shrub_Percent, Agriculture_Percent, Impervious_Percent)
```

##### Road density
```{r}
watersheds_roads <- watersheds %>%
  road_density_grabber() %>%
  select(index, road_density_km_per_km2) %>%
  st_drop_geometry()
```

#### Bind all together:
```{r}
watersheds_with_vars <- watersheds %>%
  left_join(watersheds_aspects, by = "index") %>%
  left_join(watersheds_daymet, by = "index") %>%
  left_join(pet_avg, by = "index") %>%
  left_join(watersheds_sp, by = "index") %>%
  left_join(watersheds_geology, by = "index") %>%
  left_join(watersheds_nlcd, by = "index") %>%
  left_join(watersheds_roads, by = "index") %>%
  left_join(streamcat_sim, by = "index") %>%
  st_drop_geometry()

names(watersheds_with_vars) <- tolower(names(watersheds_with_vars))

write_csv(watersheds_with_vars, "data/watersheds_with_vars.csv")
st_write(watersheds, "data/watersheds_shapefile.shp", append = FALSE)
```