---
  title: "Model_Selection"
author: "Katie Willi"
date: "2025-05-07"
output: html_document
---
  
```{r, warning=FALSE, message=FALSE}
library(tidyverse)   
library(hydroGOF)     # evaluation metrics (NSE, PBIAS, RMSE)
library(caret)        # model training and cross-validation
library(leaps)        # mlr variable selection
library(randomForest) # random forest mode
library(DT)
library(slickR)
library(broom)
library(MuMIn)
library(furrr)
```

## Data tidying

Munge all the different datasets together to get a single dataframe containing all sites, flow stats, and predictor variables:

```{r}
flow <- read_csv("data/archive/dataset_20250623_2.csv") %>%
  select(
    index, gage_used,
    all_jan_Q_mmd, all_feb_Q_mmd, all_mar_Q_mmd, all_apr_Q_mmd, all_may_Q_mmd, all_jun_Q_mmd,
    all_jul_Q_mmd, all_aug_Q_mmd, all_sep_Q_mmd, all_oct_Q_mmd, all_nov_Q_mmd, all_dec_Q_mmd,
    all_Q_ann_mm,
    all_annual_mean_max_Q_mmd, all_annual_mean_min_Q_mmd,
    all_annual_mean_Q_mmd, all_q95_Q_mmd, all_q5_Q_mmd,
    all_mean_flowdate_0.1, all_mean_flowdate_0.2, all_mean_flowdate_0.3, all_mean_flowdate_0.4,
    all_mean_flowdate_0.5, all_mean_flowdate_0.6, all_mean_flowdate_0.7, all_mean_flowdate_0.8,
    all_mean_flowdate_0.9, all_monsoon_frac,
    flood_freq_1.5_Q_mmd
  ) %>%
  rename_with(tolower)

predictors <- read_csv("data/0-FINAL-DELIVERABLES/watersheds_with_vars_20250624.csv") %>%
  select(-c(gage_used, percent_burned))

vars <- flow %>%
  inner_join(predictors, by = "index") %>%
  select(index, name, gage_used, cdwr_id, usgs_id, comid, everything()) %>%
  # These are severe outliers so we remove them:
  filter(!gage_used %in% c("9358550", "LUARMOCO"))
```


```{r}
all_flow_stats <- vars %>%
  select(all_jan_q_mmd, all_feb_q_mmd, all_mar_q_mmd, all_apr_q_mmd, all_may_q_mmd, all_jun_q_mmd,
         all_jul_q_mmd, all_aug_q_mmd, all_sep_q_mmd, all_oct_q_mmd, all_nov_q_mmd, all_dec_q_mmd,
         all_q_ann_mm,
         all_annual_mean_max_q_mmd, all_annual_mean_min_q_mmd,
         all_annual_mean_q_mmd, all_q95_q_mmd, all_q5_q_mmd,
         all_mean_flowdate_0.1, all_mean_flowdate_0.2, all_mean_flowdate_0.3, all_mean_flowdate_0.4,
         all_mean_flowdate_0.5, all_mean_flowdate_0.6, all_mean_flowdate_0.7, all_mean_flowdate_0.8,
         all_mean_flowdate_0.9, all_monsoon_frac,
         flood_freq_1.5_q_mmd) %>%
  names() %>%
  tolower()

all_predictor_vars <- names(vars)[c(37:ncol(vars))]
```

## Model development

For handling collinearity, we first remove highly correlated predictor
variables (correlation threshold of 0.85), and keep the variable that
has a stronger correlation with the target flow metric when removing one
of a pair of highly correlated predictors. This is how CSUFlow18 did it.

We then perform an "exhaustive" search to find the best combination of up to n/15 predictor variables. All y vars are square root transformed, but back-transforms predictions and observations
for evaluation metrics. The best model is selected based on the highest adjusted R-squared value.

```{r}
model_flow_metric <- function(metric_index) {
  yvar <- all_flow_stats[metric_index]
  print(paste("Predicting:", yvar))
  
  month_names <- tolower(month.abb)
  prcp_months <- paste0("prcp_", month_names)
  swe_months  <- paste0("swe_", month_names)
  monthly_vars <- c(prcp_months, swe_months)
  
  month <- str_match(yvar, "all_([a-z]{3})_q_mmd")[, 2]
  
  if (!is.na(month)) {
    month_index <- match(month, month_names)
    prior_months_index <- ((month_index - 3):month_index) %% 12
    prior_months_index[prior_months_index == 0] <- 12
    keep_months <- month_names[prior_months_index]
    
    keep_prcp <- paste0("prcp_", keep_months)
    
    swe_cutoff <- c("jan", "feb", "mar", "apr", "may", "jun")
    keep_swe_months <- intersect(keep_months, swe_cutoff)
    keep_swe <- paste0("swe_", keep_swe_months)
    
    if (!month %in% swe_cutoff) {
      keep_swe <- "swe_peak"
    }
    
    other_predictors <- setdiff(all_predictor_vars, monthly_vars)
    other_predictors <- setdiff(other_predictors, c("swe_peak", "swe_avg",
                                                    "prcp_avg_sum", "prcp_peak", "prcp_avg"))
    current_predictors <- c(keep_prcp, keep_swe, other_predictors)
  } else {
    current_predictors <- all_predictor_vars
  }
  
  allVars <- c(yvar, current_predictors)
  df_valid <- vars[, c("gage_used", allVars)] %>% na.omit()
  site_names <- df_valid$gage_used
  df_valid <- df_valid[, allVars]
  
  y <- df_valid[[yvar]]
  y_sqrt <- sqrt(y)
  allX <- df_valid %>% select(all_of(current_predictors))
  
  predictors <- allX %>% select(where(is.numeric), -c(hyd_region, dominant_aspect, dominant_geology))
  corr_matrix <- cor(predictors, use = "pairwise.complete.obs")
  corr_with_y <- abs(cor(predictors, y_sqrt, use = "pairwise.complete.obs"))
  
  vars_to_remove <- c()
  for(i in 1:(ncol(predictors) - 1)) {
    for(j in (i + 1):ncol(predictors)) {
      if(abs(corr_matrix[i, j]) >= 0.85) {
        if(corr_with_y[i] < corr_with_y[j]) {
          vars_to_remove <- c(vars_to_remove, colnames(predictors)[i])
        } else {
          vars_to_remove <- c(vars_to_remove, colnames(predictors)[j])
        }
      }
    }
  }
  
  
  predictors_corr_removed <- allX %>% select(-any_of(unique(vars_to_remove)))
  predictors_names_removed <- names(predictors_corr_removed)
  
  group_list <- list(
    hyd_region = "hyd_region",
    dominant_aspect = "dominant_aspect",
    dominant_geology = "dominant_geology"
  )
  
  numeric_vars <- setdiff(predictors_names_removed, c("hyd_region", "dominant_aspect", "dominant_geology"))
  for (v in numeric_vars) group_list[[v]] <- v
  group_list <- group_list[lengths(group_list) > 0]
  group_names <- names(group_list)
  
  all_combos <- unlist(
    lapply(1:min(as.integer(nrow(df_valid) / 15), length(group_names)),
           function(k) combn(group_names, k, simplify = FALSE)),
    recursive = FALSE
  )
  
  results <- purrr::map(all_combos, function(groups) {
    predictors_subset <- unlist(group_list[groups])
    df_X <- df_valid[, predictors_subset, drop = FALSE]
    df_X[[yvar]] <- y_sqrt
    formula <- as.formula(paste("`", yvar, "` ~ ", paste(predictors_subset, collapse = " + "), sep = ""))
    mod <- tryCatch(lm(formula, data = df_X), error = function(e) NULL)
    if (!is.null(mod)) {
      r2 <- summary(mod)$adj.r.squared
      aic <- AIC(mod)
      list(groups = groups, formula = formula, r2 = r2, aic = aic, model = mod)
    } else NULL
  })
  
  results <- compact(results)
  if (length(results) == 0) return(NULL)
  
  best_model <- results[[which.max(map_dbl(results, "r2"))]]
  katie_best_formula <- best_model$formula
  katie_final_model <- best_model$model
  
  katie_train_predictions <- predict(katie_final_model, newdata = df_valid)^2
  katie_train_obs <- df_valid[[yvar]]
  
  df_valid_sqrt <- df_valid
  df_valid_sqrt[[yvar]] <- y_sqrt
  df_valid_sqrt$site_name <- site_names
  
  cv_model <- train(katie_best_formula,
                    data = df_valid_sqrt,
                    method = "lm",
                    trControl = trainControl(method = "cv", number = 5, savePredictions = "final"),
                    metric = "RMSE")
  
  cv_pred <- cv_model$pred$pred^2
  cv_obs <- cv_model$pred$obs^2
  cv_site_names <- site_names[cv_model$pred$rowIndex]
  
  model_comparison <- data.frame(flow_stat = yvar,
                                 nobs = nrow(df_valid),
                                 model_type = c("MLR_Train", "MLR_CV"),
                                 NSE = c(NSE(katie_train_predictions, katie_train_obs), NSE(cv_pred, cv_obs)),
                                 PBIAS = c(pbias(katie_train_predictions, katie_train_obs), pbias(cv_pred, cv_obs)),
                                 R2 = c(cor(katie_train_predictions, katie_train_obs)^2, cor(cv_pred, cv_obs)^2),
                                 RMSE = c(rmse(katie_train_predictions, katie_train_obs), rmse(cv_pred, cv_obs)),
                                 CV = c(rmse(katie_train_predictions, katie_train_obs) / mean(katie_train_obs) * 100, rmse(cv_pred, cv_obs) / mean(cv_obs) * 100),
                                 vars = rep(paste(deparse(katie_best_formula), collapse = ""), 2))
  
  train_df <- data.frame(flow_stat = yvar, model_type = "MLR_", dataset = "Train",
                         site_name = site_names, observed = katie_train_obs,
                         predicted = katie_train_predictions, row_id = 1:nrow(df_valid))
  
  cv_df <- data.frame(flow_stat = yvar, model_type = "MLR_", dataset = "CV",
                      site_name = cv_site_names, observed = cv_obs,
                      predicted = cv_pred, row_id = cv_model$pred$rowIndex)
  
  list(
    model_stats = model_comparison,
    obs_pred = bind_rows(train_df, cv_df),
    model_object = list(model = katie_final_model, formula = katie_best_formula),
    yvar = yvar
  )
}

  results_list <- map(1:length(all_flow_stats), function(i) {
    res <- model_flow_metric(i)
    if (is.null(res)) return(NULL)
    yvar <- res$yvar
    # save model stats
    write_csv(res$model_stats, paste0("data/model_results_no_burn/all_model_stats_", yvar, ".csv"))
    # save observed vs predicted
    write_csv(res$obs_pred, paste0("data/model_results_no_burn/obs_pred_", yvar, ".csv"))
    return(res)
  }
  )

```

## Saving Models

```{r}
model_list <- list.files("data/model_results_no_burn/", pattern = "all_model_stats", full.names = TRUE) %>%
  map_dfr(~read_csv(.)) %>%
  filter(model_type == "MLR_Train") %>%
  mutate(flow_stat = tolower(flow_stat))

grab_models <- function(metric_index){
  
  yvar <- all_flow_stats[metric_index]
  
  print(paste("Grabbing:", yvar))
  
  current_predictors <- all_predictor_vars 
  
  allVars <- c(yvar, current_predictors)
  df_valid <- vars[, c(allVars)] %>% na.omit() 
  
  model <- model_list %>%
    filter(model_type == "MLR_Train") %>%
    filter(flow_stat == yvar) %>%
    pull(vars) 
  
  # get model formula and modify for square root transformation
  model_formula <- as.formula(model)
  
  # create new formula with sqrt transformation of y variable
  y_var_name <- as.character(model_formula)[2]   # y variable name
  predictor_terms <- as.character(model_formula)[3] %>% tolower() # predictors
  
  # modified formula with sqrt transformation:
  sqrt_formula <- as.formula(paste("sqrt(", y_var_name, ") ~", predictor_terms) %>% tolower()) 
  
  # fit the linear model with square root transformed y variable
  fitted_model <- lm(sqrt_formula, data = df_valid)
  
  filename <- paste0("data/model_obs_no_burn/model_", yvar, ".rds")
  
  # save the model as RDS for use in shiny app
  saveRDS(fitted_model, file = paste0("data/model_obs_no_burn/model_", yvar, ".rds"))
  
  return(fitted_model)
}

unique(1:n_distinct(model_list$flow_stat)) %>%
  walk(~grab_models(.)) 
```

